Analyzing changes in Canadian grocery prices

### 2024-11-14 paper/paper.qmd

## Price Difference Analysis

We further analyze `price_difference`, which captures the change between the current price (`price_current`) and the previous price (`price_old`). This variable provides insights into recent price adjustments across vendors.

@@

## Relationship Between Previous and Current Prices

We assess the relationship between previous prices (`price_old`) and current prices (`price_current`) to understand price consistency and potential price elasticity across products.

@@

## Distribution of Price Differences

To further explore pricing dynamics, we examine the distribution of `price_difference`, capturing both price increases and decreases across the dataset.

@@

>>>>

```
@fig-avg-curr-price displays the average current prices across vendors, revealing notable pricing differences. Metro and Galleria, for example, tend to have higher average prices, while No Frills and Walmart are on the lower end. These discrepancies suggest differing market positioning, with some vendors potentially catering to more price-sensitive customers.

**Implications for Consumers**: Budget-conscious consumers might prefer Walmart or No Frills for lower average prices, while those seeking specific product selections or brands might choose Metro or Galleria despite higher prices.

## Price Difference Analysis
We further analyze price_difference, which captures the change between the current price (`price_current`) and the previous price (`price_old`). This variable provides findings into recent price adjustments across vendors.

@@

```
@fig-avg-price-diff illustrates the average price differences by vendor. Retailers like Walmart and Galleria exhibit negative average price differences, suggesting a tendency to reduce prices, likely through competitive discounts. On the other hand, Loblaw shows relatively stable or small positive differences, suggesting less frequent or minor price adjustments.

**Competitive Strategy**: Vendors with negative average differences may be aiming to attract customers through regular discounts, while vendors with stable or positive differences might rely on consistent pricing for brand loyalty.

## Relationship Between Previous and Current Prices
To understand price consistency and elasticity, we assess the relationship between previous prices (`price_old`) and current prices (`price_current`).

@@

```
@fig-past-curr shows a strong positive relationship between previous prices and current prices. This pattern indicates that higher past prices are generally associated with higher current prices, implying stability in grocery pricing over time.

**Price Elasticity Findings**: The linear trend suggests low price elasticity, where previous prices predict current prices with relative stability. This may reflect supply chain consistency and retailers’ cautious approach to pricing changes.

## Distribution of Price Differences
To further explore pricing dynamics, we examine the distribution of `price_difference`, capturing both price increases and decreases across the dataset.

@@

```
@fig-dist-price (filtered to remove extreme values), shows that most price changes are small, with values clustered around zero. Both positive and negative price differences are present, indicating a balance between price increases and decreases across vendors.

**Market Stability**: The concentration around zero suggests a generally stable market, with moderate pricing adjustments. While some vendors increase prices, others offer discounts, contributing to a balanced market.
